Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-30 Manuscript/manuscript.Rmd
intro edits

# INTRODUCTION

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed the spatial organization of the human genome  [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Given their high degree of conservation between cell types and species, it has been proposed that TADs represent the fundamental building blocks of organization, giving the genome its 3D structure [@rocha2015breaking]. Studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome. It has been thought that the collection of particular functional genomic elements may contribute to the formation of TADs. Namely, the CCCTC-binding factor (CTCF) has been shown to be enriched at TAD boundaries [@ghirlando2016ctcf]. Additional transcription factors including SMC1, SMC3, and RAD21 have also been shown to colocalize near TAD boundaries [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, chromatin has been shown to exhibit certain modifications at TAD boundaries, evidenced by the frequent observation of the H3K36me3 histone mark [@dixon2012topological]. However, the growing number of publically available annotated genomic elements and there collinearity pose a significant problem for establishing methods robust enough to determine the most important epigenomic factors that contribute to TAD boudandary formation. Moreover, the plasticity of functional annotations make TAD boundary prediction a cell line specific problem [@wang2012widespread]

>>>>

# Introduction

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed the spatial organization of the human genome  [@lieberman2009comprehensive]. The genome has been shown to form tight, compact folds in order to fit inside the nucleus. This folding does not occur randomly; in fact, the genome exhibits a hierarchical 3D structure [@dixon2012topological]. Among this hierarchy are distict compartments where the genome folds and interacts with itself more frequently termed topologically associating domains (TADs). TADs have garnered much interest in recent years given that they are heavily linked to gene expression [@dixon2015chromatin]. Moreover, changing the 3D structure of the DNA, causing disruptions in the boundaries that demarcate TADs, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. It has been thought that the collection of particular functional genomic elements may contribute to the formation of TADs. Namely, the CCCTC-binding factor (CTCF) has been shown to be enriched at TAD boundaries [@ghirlando2016ctcf]. Additional transcription factors including SMC1, SMC3, and RAD21 have also been shown to colocalize near TAD boundaries [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, chromatin has been shown to exhibit certain modifications at TAD boundaries, evidenced by the frequent observation of the H3K36me3 histone mark [@dixon2012topological].

## Exhisting problems and limitations

A growing number of studies have proposed computational tools used to identify additional key molecular drivers of the formation of TAD boundaries. The classical approach is to build classification models that leverage the relationship between ChIP-seq genomic annotations and TAD boundary regions. However, the growing number of publically available annotated genomic elements, and there collinearity, pose a significant problem for establishing methods robust enough to determine the most important genomic factors that contribute to TAD boudandary formation. Likewise, it is unclear what the best method is for quantifying this relationship. Incorporating a distance measure has largely been ignored, but could offer a much more spatially descriptive feature space thereby improving predictive performance.

Another key factor in these approaches becomes resolution. Resolution determines the bin sizes of the linear genome used to enumerate contacts caused by chromatin folding. Due to high sequencing cost, most available Hi-C datasets have relatively low resolution, in the kilobase range. This is in stark contrast to ChIP-seq experiments which annotate regions of the genome at hundreds of base pairs in resolution. However, simply increasing sequencing depth to obtain higher resolution could have unintended consequences since this could introduce sparsity in the number of contacts which TAD callers are often ill-equiped to handle [@cresswell2019spectraltad, @sexton2012three]. It is unclear how varying levels of resolution affect the association between functional genomic annotations and TAD boundary regions. Increased resolution could also further exacerbate the class imbalance created from the proportionally much smaller number of TAD boundary regions compared to non-TAD boundary regions throughout the linear genome. Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing].

## Previous methods

Recently, an L1-regularized multiple logistic regression model, called *HiCfeat*, 
was developed to predict TAD boundaries using transcription factor binding sites 
on GM12878 [@mourad2016computational].
